---
title: "lab07_assignment"
format: 
  html:
    embed-resources: true
editor: visual
---

# Group Assignment Lab 7

s184343 - Lena Schlüter Nielsen

s175357 - Bolette Arildsen

s215008 - Emma Lundsbøl Wernersson

s224373 - Caroline Degner Lundsteen

s211930 - Frederik Allingham de Cordier

## Libraries

```{r}
library(tidyverse)
library(broom)
```

## Load Data

```{r}
#| message: false
biopsy <- read_csv("https://wilkelab.org/classes/SDS348/data_sets/biopsy.csv")
```

## Investigate data

The data is a breast cancer from University of Wisconsin Hospitals, Madison from Dr. William H. Wolberg. It consist of 699 patients and 10 variables. 9 attributes was scored on a scale from 1-10 and the outcome benign/malign is also present.

```{r}
str(biopsy)
```

## Data Wrangling

PCA can only be performed on numerical data, thus only numerical data is selected. Thereby the outcome-variable is removed. Since the PCA captures the variance in data, it must be scaled.\
Last PCA is performed.

```{r}
pca_fit <- biopsy |>
  select(where(is.numeric)) |>
  scale() |>
  prcomp() 
```

## Look at the data in PC coordinates

Looking at the fitted dataset firstly

```{r}
pca_fit
```

Plotting the data in PC coordinates

```{r}

pca_fit %>%
  augment(biopsy) %>% # add original dataset back in
  ggplot(aes(.fittedPC1, .fittedPC2, color = outcome)) + 
  geom_point(size = 1) +
  scale_color_manual(
    values = c(malignant = "red", benign = "green")
  ) +
  theme_classic() + background_grid()
  
```

## Look at the rotation matrix

```{r}
# extract rotation matrix
pca_fit |> 
  tidy(matrix = "rotation")
```

```{r}
# define arrow style for plotting
arrow_style <- arrow(
  angle = 20, ends = "first", type = "closed", length = grid::unit(8, "pt")
)

# plot rotation matrix
pca_fit %>%
  tidy(matrix = "rotation") |> 
  pivot_wider(names_from = "PC", names_prefix = "PC", values_from = "value") |> 
  ggplot(aes(PC1, PC2)) +
  geom_segment(xend = 0, yend = 0, arrow = arrow_style) +
  geom_text(
    aes(label = column),
    hjust = 1, nudge_x = -0.02, 
    color = "blue"
  ) +
  xlim(-1.25, .5) + ylim(-.5, 1) +
  coord_fixed() + # fix aspect ratio to 1:1
  theme_minimal_grid(12)
```

## Look at the variance explained by each P

```{r}
pca_fit %>%
  tidy(matrix = "eigenvalues")
```

```{r}
pca_fit %>%
  tidy(matrix = "eigenvalues") |> 
  ggplot(aes(PC, percent)) +
  geom_col(fill = "orange", alpha = 0.8) +
  scale_x_continuous(breaks = 1:9) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.01))
  ) +
  theme_minimal_hgrid(12)
```

By only using the first PC 65% of the variance is explained. It also becomes clear that adding additional components does not cover a significant amount of variance, and thus it could be argued that running the PCA with the single PC whilest not optimal is still doable.
